Exception Handling Bug Hazards in Android

### 2016-04-19 SPRINGER_JOURNAL/msr_journal.tex
addressing journal reviewers comments - REVIWER 3
Remove build artifacts, add Makefile
A few corrections

\institute{Roberta Coelho \at
              Federal University of Rio Grande do Norte, CIVT/UFRN. Av. Senador Salgado Filho, 3000. 
Lagoa Nova, CEP: 59.078-970. Natal/RN. Brazil \\
              Tel.: +55-84-3342-2216 \\
              \email{roberta@dimap.ufrn.br}           
           \and
          Lucas Almeida \at
             \email{lucas.almeida@ppgsc.ufrn.br} \\    
          \and    
         Georgios Gousios   \at
  \email{georgios@cs.ru.nl} \\    
 \and
         Arie van Deursen  \at
  \email{arie.vandeursen@tudelft.nl} \\      
\and 
   Christoph Treude \at
  \email{ctreude@dimap.ufrn.br} \\    
}

@@

\begin{figure*} 
\centering \includegraphics[width=.9\hsize]{overview_review.png}
\caption{Study overview.}\label{fig:overview} 
\end{figure*}

@@

@@

Table~\ref{tab:exampeswrap} presents examples of exceptions that 
were actually wrapped in this code snippet: java.lang.RuntimeException wrapped
an java.lang.OutOfMemoryError and a java.lang.StackOverflowError. 
\textcolor{red}{Such Errors result from failures detected by the runtime environment which indicate 
resource deficiencies from which the program cannot possibly recover.
Hence, `hiding" an unrecoverable condition into a runtime exception
can lead to a scenario where the developer tries to handle such exception, leading the program
to na unpredictable state. This kind of wrapping should be avoided, since represent a
serious bug hazard.}

@@

\emph{\textbf{Checked Exception wrapping an Error.}} Most of these wrappings
 were also caused by the reflection library used by applications' methods. The methods responsible for the wrappings
were also native methods written in C. Table~\ref{tab:exampeswrap} illustrates
some of these wrappings -- some of them are masking an OutOfMemoryError
into a checked exception. 
\textcolor{red}{By confronting the API user with a checked exception,
the API designer is forcing the client to handle the exceptional condition. 
But in this case the exception condition in and unrecoverable error,
which makes infeasible the task of providing a proper handler. 
This kind of wrapping is even more dangerous the previous one
and may lead to the ``exception confusion'' described next.}

@@

\textcolor{red}{As previously menitoned, the goal of this work is to obtain a thorough understanding of common exception handling \emph{bug hazards} that app developers face. In the first phase of this work, we conducted a mining study which identified common exception handling bug hazards in app development.
 In the second phase of this work, we set up an exploratory qualitative investigation and surveyed Android developers on how they perceive the bug hazards detected in the mining study. The scope of our study is GitHub---using our GHTorrent database~\cite{Gousi13}, we aimed our survey at developers from Android projects available in GitHub for which issues were mined during the first phase of this study.}

@@

\textcolor{red}{The general research question guiding our exploratory survey is twofold \emph{RQ 2: How do developers deal with the exception handling code in Android apps and what are the developers' perspectives about the main bug hazards found during the mining study?} This general research question has been broken into a set of research questions that are answered by the exploratory survey.}

@@

\textcolor{red}{Since our aim is to learn from a large number of developers, we use surveys which scale well.}
\textcolor{red}{The survey is split into two logical sections; the motaivations behind each logical section are: 
(i) questions aiming at learning from developers about the usage of the exception handling code in app development;  and 
(ii) questions focusing on getting developers percerptionds about the bug hazards detected in the first phase of this study.
The survey comprises multiple choice or Likert-scale questions and open-ended questions. The multiple choice questions are intermixed with open-ended questions to further elicit the developer's opinions. Moreover, the survey also contains Likert-scale questions to force participants to make a choice. Overall, the survey includes 13 open-ended questions, 5 Likert-scale questions and 10 multiple choice questions. The respondents could complete the survey in about 15 minutes.}

\textcolor{red}{Given the exploratory nature of our research, we used methods from Grounded Theory ~\cite{charmaz2006} to answer some of our research questions by iterating through the open-ended survey responses. The grounded theory coding used in this study consists of at two phases: (1) initial coding entails a close reading of the data and (2) later we used focused coding ~\cite{charmaz2006} to pinpoint and develop the most salient themes ~\cite{charmaz2006} in the analyzed data. Answers to different questions in the survey were coded separately. For all open-ended questions in the survey, coding was done by two researchers until saturation was reached. Disagreements were discussed and resolved as part of the coding, thus, we are unable to report level of agreement. In all cases, the first author was one of the two coders, and one of the other authors coded the data as well in close collaboration with the first author until saturation was reached. The first author then went through the remaining responses to assign codes to responses.}

@@

%~\cite{coelhoetal15}
 In the first phase of this study, described previously, we mined exception stack traces available in issues reported for several Android projects hosted on GitHub and Google Code. To ensure that our sample consists of repositories that were real Android apps, we had to inspect every project site and discard toy-programs and non-Android repositories. For each repository, we extract the developers whose emails are registered. We emailed the 1,824 developers and received 71 valid answers. \textcolor{red}{The response rate was 3.9\% -- 
although the response rate achieved in our study is low, it is inlined to similar surveys reported in the literature (e.g., ~\cite{kochhar2015understanding,joorabchi2013real,bavota2015impact,ko2007information,hindle2015topics}) whose response rates mostly varies between 2\% and 4\%.}
The majority of our respondents have more than 2 years of Java development experience (87.3\%) and of Android development (85.9\%) -- see Figure~\ref{fig:devexpertise}.

@@

%\emph{EH Best Practices}
%/////// JAVA BEST PRACTICES ////// 
As mentioned before several general guidelines have been proposed on how to
use Java exceptions. Section~\ref{sec:best}  compiled a list of some exception handling (EH) best practices that guided our mining study. In this survey the developers were also asked if they knew about and used Java EH best practices -- see Figure~\ref{fig:allquest}. Most of them said that they knew and adopted Java EH practices (68\%). From this set the most mentioned best practices are presented in Table~\ref{tab:javapractices} -- considering that one developer mentioned more than one practice. Most of the best practices mentioned were the ones cited in Section~\ref{sec:best}. For instance:\emph{``Don't throw [pure] RuntimeException (in most cases), don't catch OutOfMemoryError, etc." [D42]}.  Some of the practices mentioned, however, are not well known Java EH best practices but seem to be used by some Android developers to better deal with exceptions in the Android app context, such as the use of a crash fast approach\footnote{\url{http://www.slideshare.net/pyricau/crash-fast}} -- \emph{``Crash is good for hinting developer. Fast crash fast solve." [D51]}. \textcolor{red}{According to this approach, developers should let the application to crash as soon as an unexpected situation happens. The motivation behind this approach is that the developer will receive the crash information and fix the potential bug that caused it. On the other hand, if the developer lets the app to continue runing, after an exception had happened, the aplication may enter an inconsistent state. In such scenario, the time between the fault being exercised and the failure (the fault manifestation) increases, which may impair finding and fixing the potential bug that caused the failure.} Moreover, some respondents favor checked exceptions instead of runtime to represent the exceptional conditions in Android apps -- \emph{``RuntimeExceptions are not particularly well suited for Android's robustness, as any uncaught one would cause a crash." [D26]}; \emph{``Checked exception help, unchecked exceptions don't." [D25]}; \emph{``avoid unchecked exceptions except if the case is really abnormal and should not happen [...]"[D45]}.

@@

\textcolor{red}{This section discusses lessons learned from the mining and the survey studies. The mining study revealed a set of bug 
hazards -- such as (i) the cross-type wrappings; (ii) the abundance of null pointer problems; and (iii) the undocumented runtime 
exceptions signaled by third-party code -- which were confirmed by developers during the survey. Next subsections discuss about each of such bug hazards, presenting how the developer's perceived them and pointing to what can be done to help developers 
to deal with them.}

@@

\textcolor{red}{Another internal threat to validity of this study regards to the process adopted to 
identify the type of each exception reported on stack traces. 
To perform such type analysis we chose a version of the souce-code or bytecode
 of each reported exception, as follows: (i) for all exceptions signaled by the Android Platform 
and the Java Environment the analysis was based on Version 4.4 of Android platform (API level 19);  (ii) 
for the exceptions signaled by applications the analysis considered
the last version of the application source-code available on the Github/Google code; and (iii) 
for all exceptions signaled by third-party libraries the type analysis
considered the last bytecode version available on the app repository.
When the exception could not be found automatically or manually (based on the chosen version), 
we classified the exception as ``Undefined''.  Only 31 exceptions 
remained undefined, which occurred in 60 different exception stack traces out of 6,005 mined stack traces.
Hence, the exception type could not be identified only in 1\% of the mined flows.
There should be situations where the exception type (i.e., checked or runtime) changes from version to another. 
Altough we believe that it will not happen very often, we did not investigate such issue.
 One way to mitigate such threat would be to consider the issue date to select the most 
suitable source-code/bytecode version to perform the analysis.
However, the number of applications analyzed (approx. 6000) 
would increase the time and complexity needed to follow such improved process.}

@@

\textit{Exploratory Survey Studies.} Exploratory surveys have been used in the software
engineering context to discover the user perspective regarding a broad range of 
topics such as: assessing the developers' perceptions on productivity~\cite{meyer2014}; 
how GitHub developers use pull-requests~\cite{gousios2015}; 
how the testing culture of open-source projects can be characterized~\cite{pham2013}; 
and even how developers use Twitter~\cite{singer2014}. This kind of study is important as a way of 
better understanding the developers' behavior and hence providing recommendations and 
tools to help with specific development tasks. 
 \textcolor{red}{Some of such exploratory surveys were applied to Android developers~\cite{kochhar2015understanding,joorabchi2013real,bavota2015impact,linares2015developers}. 
Kochhar et al. ~\cite{kochhar2015understanding} conducted a survey-based study to discover the 
commonly used tools for mobile app testing as well as the problems 
faced by developers while testing the apps. The survey was sent to 3,905 emails of Android developers 
and received 83 responses (response rate of 2.13%). 
Joorabchi et al. ~\cite{joorabchi2013real} performed a survey-based study whose goal was to better understand the the main challenges 
developers face when building apps for different mobile devices. 
They interviewed 12 senior mobile developers and performed a semi-structured survey, with 188 respondents
 from the mobile development community - since the survey was shared via e-mail groups and social media websites
response rate of the survey could not be calculated. Linares-Vásquez et al. ~\cite{linares2015developers} surveyed 485 open source Android 
app and library developers (hosted on GitHub)  to understand developers’ practices for detecting and fixing 
performance bottlenecks in mobile apps.  This work emailed the survey to 24,340 developers and received 
628 responses - the response rate was 2.6\%.  
Bavota at al. ~\cite{bavota2015impact} investigated the impact of API change-proness and fault-proneness on the user ratings of Android apps. 
They surveyed developers to assess their perspective on whether such problems could be the cause for unfavorable user ratings. 
The response rate was 4\%.
A common characteristic among such works and our work is the low response rate of the surveys (between 4\% and 2\%).
None of these surveys, however, assed the how developers deal with exceptions in Android nor 
the developers' perspective regarding a set of exception handling bug hazards. In our study
we conduced the first exploratory survey focusing on exception handling issues in Android development.}

>>>>

\institute{Roberta Coelho \at
              Federal University of Rio Grande do Norte, CIVT/UFRN. Av. Senador Salgado Filho, 3000.
Lagoa Nova, CEP: 59.078-970. Natal/RN. Brazil \\
              Tel.: +55-84-3342-2216 \\
              \email{roberta@dimap.ufrn.br}
           \and
          Lucas Almeida \at
             \email{lucas.almeida@ppgsc.ufrn.br} \\
          \and
         Georgios Gousios   \at
  \email{g.gousios@cs.ru.nl} \\
 \and
         Arie van Deursen  \at
  \email{arie.vandeursen@tudelft.nl} \\
\and
   Christoph Treude \at
  \email{ctreude@dimap.ufrn.br} \\
}

@@

\begin{figure*}
%\centering \includegraphics[width=.9\hsize]{overview_review.png}
\centering \includegraphics[width=1\hsize]{overview_review.png}
\caption{Study overview.}\label{fig:overview}
\end{figure*}

@@

%TABLE 5 WAS HERE

@@

Table~\ref{tab:exampeswrap} presents examples of exceptions that
were actually wrapped in this code snippet: java.lang.RuntimeException wrapped
an java.lang.OutOfMemoryError and a java.lang.StackOverflowError.
\textcolor{red}{Such Errors result from failures detected by the runtime environment which indicate
resource deficiencies from which the program cannot possibly recover.
Hence, ``hiding" an unrecoverable condition into a runtime exception
can lead to a scenario where the developer tries to handle such an exception, leading the program
to an unpredictable state. This kind of wrapping should be avoided, since represent a
serious bug hazard.}

@@

\emph{\textbf{Checked Exception wrapping an Error.}} Most of these wrappings
 were also caused by the reflection library used by applications' methods. The methods responsible for the wrappings
were also native methods written in C. Table~\ref{tab:exampeswrap} illustrates
some of these wrappings --- some of them are masking an OutOfMemoryError
into a checked exception.
\textcolor{red}{By confronting the API user with a checked exception,
the API designer is forcing the client to handle the exceptional condition.
In this case, the exception condition masks an unrecoverable error,
which makes the task of providing a proper handler
infeasible.\footnote{TODO: Please note that Errors are not supposed to be
caught. The main problem here is that the program continues with e.g. no
memory available.}
This kind of wrapping is even more dangerous the previous one
and may lead to the ``exception confusion'' described next.}

@@

\textcolor{red}{As previously mentioned, the goal of this work is to obtain a thorough understanding of common exception handling \emph{bug hazards} that app developers face. In the first phase of this work, we conducted a mining study which identified common exception handling bug hazards in app development.
 In the second phase of this work, we set up an exploratory qualitative investigation and surveyed Android developers on how they perceive the bug hazards detected in the mining study. The scope of our study is GitHub---using our GHTorrent database~\cite{Gousi13}, we aimed our survey at developers from Android projects available in GitHub for which issues were mined during the first phase of this study.}

@@

\textcolor{red}{The overall research question guiding our exploratory survey is
the following:
\emph{RQ 2: How do developers deal with the exception handling code in Android apps and what are the developers' perspectives about the main bug hazards found during the mining study?} This general research question has been broken into a set of research questions that are answered by the exploratory survey.}

@@

\textcolor{red}{Since our aim is to learn from a large number of developers, we
use surveys which are known to scale well.}
\textcolor{red}{The survey is split into two logical sections; the motivations behind each logical section are:
(i) questions aiming at learning from developers about the usage of the exception handling code in app development;  and
(ii) questions focusing on getting developers perceptions about the bug hazards detected in the first phase of this study.
The survey comprises multiple choice or Likert-scale questions and open-ended questions. The multiple choice questions are intermixed with open-ended questions to further elicit the developer's opinions. Moreover, the survey also contains Likert-scale questions to force participants to make a choice. Overall, the survey includes 13 open-ended questions, 5 Likert-scale questions and 10 multiple choice questions. The respondents could complete the survey in about 15 minutes.}

\textcolor{red}{Given~\footnote{TODO: I (GG) would put this paragraph before
the previous one} the exploratory nature of our research, we used methods from Grounded Theory ~\cite{charmaz2006} to answer some of our research questions by iterating through the open-ended survey responses. The grounded theory coding used in this study consists of at two phases: (1) initial coding entails a close reading of the data and (2) later we used focused coding ~\cite{charmaz2006} to pinpoint and develop the most salient themes ~\cite{charmaz2006} in the analyzed data. Answers to different questions in the survey were coded separately. For all open-ended questions in the survey, coding was done by two researchers until saturation was reached. Disagreements were discussed and resolved as part of the coding, thus, we are unable to report level of agreement. In all cases, the first author was one of the two coders, and one of the other authors coded the data as well in close collaboration with the first author until saturation was reached. The first author then went through the remaining responses to assign codes to responses.}

@@

%~\cite{coelhoetal15}
 In the first phase of this study, described previously, we mined exception stack traces available in issues reported for several Android projects hosted on GitHub and Google Code. To ensure that our sample consists of repositories that were real Android apps, we had to inspect every project site and discard toy-programs and non-Android repositories. For each repository, we extract the developers whose emails are registered. We emailed the 1,824 developers and received 71 valid answers. \textcolor{red}{The response rate was 3.9\% --
although the response rate achieved in our study is low, it is in line with similar surveys reported in the literature (e.g., ~\cite{kochhar2015understanding,joorabchi2013real,bavota2015impact,ko2007information,hindle2015topics}) whose response rates mostly varies between 2\% and 4\%.}
The majority of our respondents have more than 2 years of Java development experience (87.3\%) and of Android development (85.9\%) -- see Figure~\ref{fig:devexpertise}.

@@

%\emph{EH Best Practices}
%/////// JAVA BEST PRACTICES //////
As mentioned before several general guidelines have been proposed on how to
use Java exceptions. Section~\ref{sec:best}  compiled a list of some exception
handling (EH) best practices that guided our mining study. In this survey the
developers were also asked if they knew about and used Java EH best practices --
see Figure~\ref{fig:allquest}. Most of them said that they knew and adopted Java
EH practices (68\%). From this set the most mentioned best practices are
presented in Table~\ref{tab:javapractices} -- considering that one developer
mentioned more than one practice. Most of the best practices mentioned were the
ones cited in Section~\ref{sec:best}. For instance:\emph{``Don't throw [pure]
RuntimeException (in most cases), don't catch OutOfMemoryError, etc." [D42]}.
Some of the practices mentioned, however, are not well known Java EH best
practices but seem to be used by some Android developers to better deal with
exceptions in the Android app context, such as the use of a crash fast
approach\footnote{\url{http://www.slideshare.net/pyricau/crash-fast}} --
\emph{``Crash is good for hinting developer. Fast crash fast solve." [D51]}.
\textcolor{red}{According to this approach, developers should let the
application to crash as soon as an unexpected situation happens. The motivation
behind this approach is that the developer will receive the crash information
and fix the potential bug that caused it. On the other hand, if the developer
lets the app to continue running, after an exception had happened, the
application may enter an inconsistent state. In such scenario, the time between the fault being exercised and the failure (the fault manifestation) increases, which may impair finding and fixing the potential bug that caused the failure.} Moreover, some respondents favor checked exceptions instead of runtime to represent the exceptional conditions in Android apps -- \emph{``RuntimeExceptions are not particularly well suited for Android's robustness, as any uncaught one would cause a crash." [D26]}; \emph{``Checked exception help, unchecked exceptions don't." [D25]}; \emph{``avoid unchecked exceptions except if the case is really abnormal and should not happen [...]"[D45]}.

@@

\textcolor{red}{This section discusses the lessons learned from the mining and the survey studies. The mining study revealed a set of bug
hazards -- such as (i) the cross-type wrappings; (ii) the abundance of null pointer problems; and (iii) the undocumented runtime
exceptions signaled by third-party code, which were confirmed by developers
during the survey. Here, we discuss the identified bug hazards, we present how
the developers perceived them and point out at what developers can do in
order to deal with them.}

@@

\textcolor{red}{The process to
identify the type of exceptions reported in stack traces may not be completely
accurate.
Specifically, we performed the selection of the version of the exception source code or bytecode
to analyze, as follows: (i) for all exceptions signaled by the Android Platform
and the Java Environment the analysis was based on Version 4.4 of Android platform (API level 19);  (ii)
for the exceptions signaled by applications, the analysis considered
the last version of the application source-code available on the Github/Google code; and (iii)
for all exceptions signaled by third-party libraries the type analysis
considered the latest bytecode version available on the app repository.
When the exception could not be found automatically or manually (based on the chosen version),
we classified the exception as ``Undefined''.  Only 31 exceptions
remained undefined, which occurred in 60 different exception stack traces out of 6,005 mined stack traces.
Hence, the exception type could not be accurately identified in 1\% of the mined
exception flows.
There should be situations where the exception type (i.e., checked or runtime) changes from version to another.
Although we believe that it will not happen very often, we did not investigate
this issue.}

% One way to mitigate such a threat would be to consider the issue date to select the most
%suitable source-code/bytecode version to perform the analysis.
%However, the number of applications analyzed (approx. 6000)
%would increase the time and complexity needed to follow such improved process.}

@@

\textit{Exploratory Survey Studies.} Exploratory surveys have been used in the software
engineering context to discover the user perspective regarding a broad range of
topics such as: assessing the developers' perceptions on productivity~\cite{meyer2014};
how GitHub developers use pull-requests~\cite{gousios2015};
how the testing culture of open-source projects can be characterized~\cite{pham2013};
and even how developers use Twitter~\cite{singer2014}. This kind of study is important as a way of
better understanding the developers' behavior and hence providing recommendations and
tools to help with specific development tasks.

\textcolor{red}{Exploratory surveys have also targeted Android developers~\cite{kochhar2015understanding,joorabchi2013real,bavota2015impact,linares2015developers}.
Kochhar et al. ~\cite{kochhar2015understanding} conducted a survey-based study to discover the
commonly used tools for mobile app testing as well as the problems
faced by developers while testing the apps. The survey was sent to 3,905 emails of Android developers
and received 83 responses (response rate of 2.13\%).
Joorabchi et al. ~\cite{joorabchi2013real} performed a survey-based study whose goal was to better understand the main challenges
developers face when building apps for different mobile devices.
They interviewed 12 senior mobile developers and performed a semi-structured survey, with 188 respondents
 from the mobile development community, since the survey was shared via e-mail groups and social media websites
response rate of the survey could not be calculated. Linares-Vásquez et al. ~\cite{linares2015developers} surveyed 485 open source Android
app and library developers (hosted on GitHub)  to understand developers’ practices for detecting and fixing
performance bottlenecks in mobile apps.  This work emailed the survey to 24,340 developers and received
628 responses - the response rate was 2.6\%.
Bavota at al. ~\cite{bavota2015impact} investigated the impact of API change-proness and fault-proneness on the user ratings of Android apps.
They surveyed developers to assess their perspective on whether such problems could be the cause for unfavorable user ratings.
The response rate was 4\%.
A common characteristic among such works and our work is the low response rate of the surveys (between 4\% and 2\%).
None of these surveys, however, assessed how developers deal with exceptions in Android nor
the developers' perspective regarding a set of exception handling bug hazards.
In our study,
we conduced the first exploratory survey focusing on exception handling issues in Android development.}
